%% paper_blind.tex
%% Blind manuscript for double-blind peer review
%% Submitted to: Systems & Control Letters
%% Author information and code repository omitted for review.

¥documentclass[final,3p,times]{elsarticle}

¥usepackage{amsmath,amssymb,amsthm}
¥usepackage{booktabs}
¥usepackage{microtype}
¥usepackage[colorlinks=true,citecolor=blue,urlcolor=blue,linkcolor=black]{hyperref}

¥input{outputs/data/paper_macros.tex}

¥newtheorem{theorem}{Theorem}
¥newtheorem{proposition}[theorem]{Proposition}
¥newtheorem{corollary}[theorem]{Corollary}
¥newtheorem{definition}[theorem]{Definition}
¥newtheorem{remark}[theorem]{Remark}

¥newcommand{¥Gr}{¥mathrm{Gr}}
¥newcommand{¥tr}{¥mathrm{tr}}
¥newcommand{¥R}{¥mathbb{R}}
¥renewcommand{¥qedsymbol}{$¥blacksquare$}

¥journal{Systems ¥& Control Letters}

¥begin{document}

¥begin{frontmatter}

¥title{Task-Dependent Optimal Linear Observers in
       Linear Gaussian Dynamical Systems}

%% Author information omitted for double-blind peer review.

¥begin{abstract}
For a stationary linear Gaussian dynamical system, the rank-$r$ linear
observer minimizing squared-error prediction loss for target $Cx_{t+¥tau}$
has row space equal to the top-$r$ eigenspace of the target information
matrix $M_{C,¥tau} = ¥Sigma^{1/2}(A^¥tau)^¥top C^¥top CA^¥tau¥Sigma^{1/2}$,
uniquely up to eigenspace rotation.
Two targets with distinct dominant eigenspaces admit no common optimal
rank-$r$ observer.
A common optimal observer exists if and only if the target family is
¥emph{$r$-coherent}: all information matrices share a common top-$r$
eigenspace.
This condition is exact, not asymptotic or approximate.
Numerical illustrations at $n=8$, $r=2$ confirm a target-dependent case
($¥Gamma=¥GammaCondA>0$) and an $r$-coherent construction, both consistent
with the theorem.
¥end{abstract}

¥begin{keyword}
linear Gaussian dynamical systems ¥sep low-rank linear observers ¥sep
optimal observation ¥sep Grassmannian ¥sep Ky Fan theorem
¥MSC[2020] 93B07 ¥sep 93E10 ¥sep 15A18
¥end{keyword}

¥end{frontmatter}

% -------------------------------------------------------------------------
¥section{Introduction}
% -------------------------------------------------------------------------

A central question in observer design for linear dynamical systems is whether
a single rank-$r$ linear observer can be simultaneously optimal for predicting
multiple output channels $¥{C_k x_{t+¥tau_k}¥}$, or whether the optimal
observer depends on the specific prediction target and horizon.
For finite-dimensional linear Gaussian dynamical systems (LGDS) at
stationarity, this paper gives an exact answer: the optimal rank-$r$ observer
for predicting $Cx_{t+¥tau}$ is the top-$r$ eigenspace of the target
information matrix $M_{C,¥tau}$, and a common optimal observer across a family
of targets exists if and only if all information matrices share a common
dominant eigenspace.

This paper is not about dynamic observer synthesis in the Luenberger--Kalman
sense ¥cite{Luenberger1971,OReilly1983,Kailath1980}: the observer is a fixed rank-$r$
linear map $¥phi_U(x)=U¥Sigma^{-1/2}x$, and performance is squared-error
prediction loss for a specified target $Cx_{t+¥tau}$ at stationarity.
The static case ($A=I$, $¥tau=0$) recovers standard principal-subspace
selection, but the main object here is the interaction of system dynamics,
prediction horizon $¥tau$, and observer rank through the information matrix
$M_{C,¥tau}$.

Reduced representations for linear systems have been studied via principal
components of Gramians ¥cite{Moore1981}, balanced truncation, and related
model-reduction methods ¥cite{Antoulas2005}.
The observability Gramian accumulates contributions from all prediction
horizons and all output directions; balanced truncation minimises
approximation error in $H_2$/$H_¥infty$ system norms rather than
squared-error prediction loss for a specified $(C,¥tau)$.
Sensor-selection and output-design methods target estimation or
reconstruction quality for a given objective, typically without
characterising when a fixed low-rank map is simultaneously optimal
across an entire task family.
The present note asks precisely that structural question: given distinct
prediction targets with information matrices $¥{M_{C_k,¥tau_k}¥}$, when
is one rank-$r$ observer simultaneously optimal for all of them, and when
is target-dependent selection unavoidable?

The Grassmannian $¥Gr(r,n)$ is the natural parameter space for rank-$r$
linear observers ¥cite{Edelman1998}, and observer quality---the trace
$¥tr(U M_{C,¥tau} U^¥top)$---is a smooth function on this manifold.
Optimal observer selection reduces to trace maximisation on $¥Gr(r,n)$,
governed by the Ky Fan variational principle ¥cite{Fan1949}: the maximum
is achieved by the leading eigenspace of $M_{C,¥tau}$, recovering
principal-subspace selection ¥cite{Hotelling1933} in the static ($A=I$)
case within the classical linear state-space framework ¥cite{Kailath1980}.

The Ky Fan variational principle and principal-subspace optimality are
classical ¥cite{Fan1949,Hotelling1933}.
The single-target result (Proposition‾¥ref{prop:opt}) applies Ky Fan to the
target information matrix $M_{C,¥tau}$, which identifies which directions of
state space bear on a given prediction task.
What is not resolved by classical eigenspace theory is the task-family
question: when does a single rank-$r$ observer remain exactly optimal across
a family of targets with possibly distinct information matrices?
The structural characterization, given in Corollary‾¥ref{cor:coherence}, is
$r$-coherence: its non-trivial content is the backward direction, which
shows that any commonly optimal observer must be the unique top-$r$ eigenspace
of each individual $M_k$, forcing all information matrices to share a common
dominant eigenspace.
The regret gap $¥Gamma$ provides a computable certificate of unavoidable
suboptimality when $r$-coherence fails.

The result is scoped to finite-dimensional LGDS, fixed-rank linear observers,
and squared-error prediction targets.

% -------------------------------------------------------------------------
¥section{System and Observer Family}
% -------------------------------------------------------------------------

¥paragraph{Dynamics.}
Let
¥begin{equation}
  x_{t+1} = A x_t + ¥xi_t, ¥quad ¥xi_t ¥sim ¥mathcal{N}(0, Q),
  ¥quad x_t ¥in ¥R^n,
¥end{equation}
with $A$ stable (spectral radius $<1$) and $Q ¥succ 0$.
The unique stationary covariance $¥Sigma$ satisfies
$A ¥Sigma A^¥top + Q = ¥Sigma$ and is strictly positive definite
¥cite{Kailath1980}; $¥Sigma^{1/2}$ denotes its symmetric positive-definite
square root.

¥paragraph{Observer family.}
An observer of rank $r$ ($1 ¥le r < n$) is parametrised by a
row-orthonormal matrix $U ¥in ¥R^{r ¥times n}$, $U U^¥top = I_r$:
¥begin{equation}
  ¥phi_U(x) = U ¥Sigma^{-1/2} x.
¥end{equation}
Pre-whitening by $¥Sigma^{-1/2}$ places all observers on a common isotropic
footing; in the static case ($A=I$, $¥tau=0$) this reduces to principal
component analysis ¥cite{Hotelling1933}.
Two observers differing only by an orthogonal rotation of their row spaces are
equivalent; the observer space is the Grassmannian
$¥Gr(r,n)$, a smooth compact Riemannian manifold of dimension $r(n-r)$,
with metric and geodesic structure from $O(n)$ ¥cite{Edelman1998}.

¥paragraph{Target family.}
A prediction target is a pair $(C, ¥tau)$ with $C ¥in ¥R^{p ¥times n}$ and
integer $¥tau ¥ge 1$:
¥begin{equation}
  y_{C,¥tau} = C x_{t+¥tau}.
¥end{equation}
Performance is total mean squared prediction error.
A target family $¥mathcal{T} = ¥{(C_k, ¥tau_k)¥}$ is any finite collection.

% -------------------------------------------------------------------------
¥section{Theorem}
% -------------------------------------------------------------------------

¥subsection{Exact Bayes risk formula}

Given observation $z = U ¥Sigma^{-1/2} x_t$, the MMSE predictor of
$Cx_{t+¥tau}$ is linear by Gaussian conjugacy.
At stationarity, $x_{t+¥tau} = A^¥tau x_t + w_¥tau$ where $w_¥tau$
is the accumulated process noise independent of $x_t$, so
$E[x_{t+¥tau}x_t^¥top] = A^¥tau¥Sigma$.
It follows that
$¥mathrm{Cov}(Cx_{t+¥tau},¥,z^¥top)
 = CA^¥tau¥Sigma¥cdot¥Sigma^{-1/2}U^¥top = CA^¥tau¥Sigma^{1/2}U^¥top$
and $¥mathrm{Cov}(z,z^¥top) = U¥Sigma^{-1/2}¥Sigma¥Sigma^{-1/2}U^¥top = I_r$.
Applying the Gaussian linear MMSE formula
$¥mathcal{R}(U;¥,C,¥tau)
 = ¥mathrm{tr}(¥mathrm{Var}(Cx_{t+¥tau}))
 - ¥mathrm{tr}¥!¥bigl(¥mathrm{Cov}(Cx_{t+¥tau},z^¥top)
                       ¥mathrm{Cov}(z,z^¥top)^{-1}
                       ¥mathrm{Cov}(z,Cx_{t+¥tau})¥bigr)$
gives
¥begin{equation}
  ¥mathcal{R}(U;¥,C,¥tau)
    = ¥tr(C¥Sigma C^¥top) - ¥tr¥!¥bigl(U¥,M_{C,¥tau}¥,U^¥top¥bigr),
¥end{equation}
where the ¥emph{target information matrix} is
¥begin{equation}
  M_{C,¥tau}
    = ¥Sigma^{1/2}(A^¥tau)^¥top C^¥top C A^¥tau ¥Sigma^{1/2}
    ¥;¥in¥; ¥R^{n¥times n}.
¥end{equation}
$M_{C,¥tau}$ is symmetric positive semidefinite of rank at most $¥min(p,n)$.
The formula is exact.

¥begin{remark}
In the static case ($A=I$, $¥tau=0$),
$M_{C,0}=¥Sigma^{1/2}C^¥top C¥Sigma^{1/2}$,
which differs from the state covariance $¥Sigma$ used by PCA unless
$C=I_n$.
For $C=e_k^¥top$ (predicting one component), $M_{C,0}$ is rank-$1$ with
eigenvector $e_k$ in the whitened basis: the optimal rank-$1$ observer
captures the $k$-th whitened direction regardless of the state variance in
other components, whereas PCA selects the maximum-variance direction.
In the dynamic case, the rotation by $(A^¥tau)^¥top$ in $M_{C,¥tau}$
shifts the dominant eigenspace with‾$¥tau$: the same target $C$ admits a
horizon-dependent optimal observer.
¥end{remark}

¥subsection{Top-eigenspace optimality}

¥begin{proposition}¥label{prop:opt}
For fixed rank $r$, the observer minimising $¥mathcal{R}(U;¥,C,¥tau)$ over
$¥Gr(r,n)$ has row space equal to the top-$r$ eigenspace of $M_{C,¥tau}$.
The minimum risk is
¥begin{equation}
  ¥mathcal{R}^*
    = ¥tr(C¥Sigma C^¥top) - ¥sum_{i=1}^{r} ¥lambda_i(M_{C,¥tau}),
¥end{equation}
where $¥lambda_1 ¥ge ¥cdots ¥ge ¥lambda_n$ are the eigenvalues of $M_{C,¥tau}$.
Uniqueness up to rotation holds when $¥lambda_r > ¥lambda_{r+1}$.
¥end{proposition}

Degeneracy at the gap ($¥lambda_r = ¥lambda_{r+1}$) yields a continuum of
optimal observers spanning the degenerate eigenspace; in practice, a generic
target $(C,¥tau)$ produces a strict spectral gap, making degeneracy
non-generic for fixed $(A,¥Sigma)$.

¥begin{proof}
Minimising $¥mathcal{R}$ is equivalent to maximising
$¥tr(U M_{C,¥tau} U^¥top)$ subject to $UU^¥top = I_r$.
By the Ky Fan variational principle ¥cite{Fan1949,HornJohnson2013}
the maximum equals
$¥sum_{i=1}^r ¥lambda_i(M_{C,¥tau})$, achieved exactly when the row space of
$U$ spans the corresponding eigenspace.
¥end{proof}

The ¥emph{task regret} of observer $U$ for target $(C,¥tau)$ is the excess
Bayes risk:
¥begin{equation}¥label{eq:regret}
  ¥mathrm{Reg}(U;¥,C,¥tau)
  ¥;:=¥;
  ¥mathcal{R}(U;¥,C,¥tau) - ¥mathcal{R}^*(C,¥tau)
  ¥;=¥;
  ¥sum_{i=1}^r¥lambda_i(M_{C,¥tau})-¥tr(U¥,M_{C,¥tau}¥,U^¥top)
  ¥;¥ge¥; 0.
¥end{equation}

¥begin{corollary}[Regret--angle bound]¥label{cor:regretbound}
Let $M=M_{C,¥tau}$ with eigenvalues $¥lambda_1¥ge¥cdots¥ge¥lambda_n¥ge 0$
and top-$r$ eigenspace $U^*$.
For any $U¥in¥Gr(r,n)$:
¥begin{equation}¥label{eq:regretbound}
  (¥lambda_r-¥lambda_{r+1})¥,¥|¥sin¥Theta¥|_F^2
  ¥;¥le¥;
  ¥mathrm{Reg}(U;¥,C,¥tau)
  ¥;¥le¥;
  (¥lambda_1-¥lambda_n)¥,¥|¥sin¥Theta¥|_F^2,
¥end{equation}
where $¥|¥sin¥Theta¥|_F^2:=¥|¥sin¥Theta(U,U^*)¥|_F^2$
¥textup{¥cite{BjorckGolub1973}} is the squared Frobenius norm of the
principal-angle sines between $¥mathrm{row}(U)$ and $¥mathrm{row}(U^*)$.
¥end{corollary}

¥begin{proof}
Let $p_i:=¥sum_{k=1}^r(v_i^¥top u_k)^2$ where $v_i$ are eigenvectors of
$M$ and $u_k$ are orthonormal rows of $U$.
Then $¥mathrm{Reg}(U)=¥sum_{i¥le r}¥lambda_i(1-p_i)-¥sum_{i>r}¥lambda_ip_i$
and $¥|¥sin¥Theta¥|_F^2=¥sum_{i¥le r}(1-p_i)=¥sum_{i>r}p_i$
¥cite{BjorckGolub1973}.
Centering at‾$¥lambda_{r+1}$:
¥[
  ¥mathrm{Reg}(U)
  =¥underbrace{¥textstyle¥sum_{i¥le r}(¥lambda_i-¥lambda_{r+1})(1-p_i)}_{¥ge 0}
  +¥underbrace{¥textstyle¥sum_{i>r}(¥lambda_{r+1}-¥lambda_i)¥,p_i}_{¥ge 0}.
¥]
¥emph{Lower bound}: drop the second sum; bound $¥lambda_i-¥lambda_{r+1}
¥ge¥lambda_r-¥lambda_{r+1}$ for $i¥le r$.
¥emph{Upper bound}: bound the first coefficient by $¥lambda_1-¥lambda_{r+1}$
and the second by $¥lambda_{r+1}-¥lambda_n$; the two $¥|¥sin¥Theta¥|_F^2$
factors sum as written.
¥end{proof}

¥subsection{Target-dependent observer selection}

¥begin{corollary}¥label{cor:noglobal}
If targets $(C_1,¥tau_1)$ and $(C_2,¥tau_2)$ have information matrices
$M_1$ and $M_2$ with distinct top-$r$ eigenspaces, no single rank-$r$ observer
is simultaneously optimal for both.
¥end{corollary}

¥begin{proof}
Optimality for target $k$ requires the row space of $U$ to equal the top-$r$
eigenspace of $M_k$.
Distinct eigenspaces cannot be spanned by the same $U$.
¥end{proof}

The corollary extends to any finite target family: if $¥{M_k¥}$ do not share
a common top-$r$ eigenspace, no common optimal rank-$r$ observer exists.

¥subsection{¥texorpdfstring{$r$}{r}-coherence: the exact exception}

¥begin{definition}¥label{def:rcoh}
A target family $¥mathcal{T}$ is ¥emph{$r$-coherent} with respect to
$(A,¥Sigma)$ if all matrices $¥{M_{C,¥tau}:(C,¥tau)¥in¥mathcal{T}¥}$
share a common top-$r$ eigenspace $V^*$.
¥end{definition}

¥begin{corollary}¥label{cor:coherence}
A common optimal rank-$r$ observer exists if and only if $¥mathcal{T}$ is
$r$-coherent.
¥end{corollary}

¥begin{proof}
(¥emph{If})¥enspace If $¥mathcal{T}$ is $r$-coherent with shared top-$r$
eigenspace $V^*$, the observer with row space $V^*$ minimises
$¥mathcal{R}(¥cdot;¥,C_k,¥tau_k)$ for each $k$ by Proposition‾¥ref{prop:opt}.
(¥emph{Only if})¥enspace If $U$ is simultaneously optimal for all $k$, then
by Proposition‾¥ref{prop:opt}'s uniqueness (holding when
$¥lambda_r(M_k)>¥lambda_{r+1}(M_k)$ for all $k$), the row space of $U$
must equal the top-$r$ eigenspace of every $M_k$; hence all $M_k$ share
that eigenspace and $¥mathcal{T}$ is $r$-coherent.
¥end{proof}

$r$-coherence requires all targets to be functions of a common $r$-dimensional
sufficient statistic across all prediction horizons---a strong alignment
between the system dynamics and the task family.
Outside $r$-coherence, the optimal observer is target-dependent.

¥begin{remark}
$r$-coherence is an algebraic equality between eigenspaces, not an
approximate or asymptotic condition.
It is the exact condition for a common optimal observer to exist.
¥end{remark}

¥begin{remark}
$r$-coherence is strictly weaker than simultaneous diagonalisability of
$¥{M_k¥}$: only the dominant $r$-dimensional eigenspace must coincide,
not the full eigenbasis.
Simultaneous diagonalisability (equivalently, pairwise commutativity
$[M_i,M_j]=0$) implies $r$-coherence for every $r$, but not conversely.
For generic $(A,¥Sigma)$ and target families, the top-$r$ eigenspaces of
distinct $M_k$ are distinct, so $r$-coherence is a non-generic condition
absent structural alignment between the task family and the system.
¥end{remark}

¥subsection{Minimax shared observer and approximate coherence}
¥label{sec:minimax}

For a task family $¥mathcal{T}=¥{(C_a,¥tau_a)¥}_{a=1}^m$, define the
¥emph{minimax shared observer} and ¥emph{minimax regret gap}:
¥begin{equation}¥label{eq:mm}
  U_{¥mathrm{mm}}
  = ¥arg¥min_{U¥in¥Gr(r,n)}¥max_a¥,¥mathrm{Reg}(U;¥,C_a,¥tau_a),
  ¥qquad
  ¥Gamma
  := ¥min_{U¥in¥Gr(r,n)}¥max_a¥,¥mathrm{Reg}(U;¥,C_a,¥tau_a).
¥end{equation}
$U_{¥mathrm{mm}}$ is the best shared rank-$r$ observer in the worst-case sense;
$¥Gamma¥ge 0$ is the unavoidable worst-case cost of any single observer.

¥begin{corollary}¥label{cor:gammazero}
$¥Gamma=0$ if and only if $¥mathcal{T}$ is $r$-coherent.
¥end{corollary}

¥begin{proof}
(¥emph{If})¥enspace If $r$-coherent with shared eigenspace $V^*$:
$¥mathrm{Reg}(V^*;¥,C_a,¥tau_a)=0$ for all $a$ by
Proposition‾¥ref{prop:opt}, so $¥Gamma=0$.
(¥emph{Only if})¥enspace If $¥Gamma=0$, some $U_{¥mathrm{mm}}$ achieves
$¥mathrm{Reg}(U_{¥mathrm{mm}};¥,C_a,¥tau_a)=0$ for all $a$, so
$U_{¥mathrm{mm}}$ is optimal for every task; the only-if direction of
Corollary‾¥ref{cor:coherence} then gives $r$-coherence.
¥end{proof}

¥begin{corollary}[Approximate coherence]¥label{cor:approxcoh}
If there exists $U¥in¥Gr(r,n)$ with
$¥|¥sin¥Theta(U,U_a^*)¥|_F^2¥le¥varepsilon^2$ for all $a$, then
$¥Gamma¥le¥Lambda¥,¥varepsilon^2$,
where $¥Lambda:=¥max_a[¥lambda_1(M_a)-¥lambda_n(M_a)]$.
¥end{corollary}

¥begin{proof}
Apply the upper bound of Corollary‾¥ref{cor:regretbound} to $U$ for each
task and take the maximum over‾$a$.
¥end{proof}

% -------------------------------------------------------------------------
¥section{Numerical Illustrations}
% -------------------------------------------------------------------------

The theorem is analytically exact under its stated assumptions.
We illustrate it numerically with two finite-dimensional instances:
a target-dependent case and an $r$-coherent case.
All reported quantities are computed exactly from the analytic Bayes risk
formula and numerical optimization on the Grassmannian.

¥paragraph{Setup.}
State dimension $n=8$, observer rank $r=2$.
The model is a finite-dimensional linear Gaussian dynamical system
¥cite{Kalman1960}.
$A = VDV^{-1}$ with eigenvalues drawn from $U[0.3,0.85]$, random
non-orthogonal $V$ (seed 100), $Q = 0.1¥,I_8$.
$¥Sigma$ is computed from the discrete Lyapunov equation; $¥Sigma¥succ 0$
is confirmed to machine precision.
All reported quantities are computed exactly from the analytic Bayes risk
formula, eigenspace calculations, and repeated projected optimization over
$¥Gr(2,8)$ (300 random initializations of projected L-BFGS-B, gradient-norm
threshold $10^{-12}$, function-value tolerance $10^{-15}$).

¥paragraph{Condition A: target-dependent instance.}
{¥sloppy
Targets $T_1 = ([e_1^¥top;¥,e_2^¥top],¥,1)$ and
$T_2 = ([e_5^¥top;¥,e_6^¥top],¥,5)$.¥par}
With non-normal $A$, $A^1$ and $A^5$ rotate state-space directions
differently; the principal angles ¥cite{BjorckGolub1973} between the top-2
eigenspaces of $M_1$ and $M_2$ are $¥CondAAngleOne^¥circ$ and
$¥CondAAngleTwo^¥circ$, confirming that no single rank-2 observer is
simultaneously optimal.

The task regret ¥eqref{eq:regret} and minimax gap $¥Gamma$
¥eqref{eq:mm} are computed from the exact Bayes risk formula.
By Corollary‾¥ref{cor:gammazero}, $¥Gamma>0$ certifies non-$r$-coherence.
The numerically computed minimax observer $U_{¥mathrm{joint}}$
achieves $¥Gamma=¥GammaCondA>0$; balanced regrets on both targets
are a necessary condition for a minimax solution, consistent with the
global minimum having been attained.
Table‾¥ref{tab:condA} gives the full regret structure‾¥eqref{eq:regret}.

¥begin{table}[tbp]
¥caption{Regret matrix, Condition A ($n=8$, $r=2$, seed 100).
All values from the exact analytic Bayes risk formula.}
¥label{tab:condA}
¥begin{center}
¥begin{tabular}{lcc}
¥toprule
Observer & Regret on $T_1$ & Regret on $T_2$ ¥¥
¥midrule
$U^*_1$ (optimal for $T_1$) & ¥CondARegretUoneTa & ¥CondARegretUoneTb ¥¥
$U^*_2$ (optimal for $T_2$) & ¥CondARegretUtwoTa & ¥CondARegretUtwoTb ¥¥
$U_{¥mathrm{joint}}$ (compromise) & ¥CondARegretUjointTa & ¥CondARegretUjointTb ¥¥
¥bottomrule
¥end{tabular}
¥end{center}
¥end{table}

¥paragraph{Condition B: structurally coherent example.}
A target family is constructed so that all $M_k$ share the same top-2
eigenspace by design: normal $A_{¥mathrm{sym}} = VDV^¥top$ with orthogonal
$V$, and a common $C$ whose rows are the top-2 eigenvectors of $A$, varying
only the prediction horizon $¥tau ¥in ¥{1,2,3,5¥}$.
Under this construction $M_{C,¥tau}$ has the same dominant 2-dimensional
eigenspace for all $¥tau$, so the four optimal observers are identical by the
theorem.
The maximum pairwise principal angle across all four is
$¥CondBMaxAngleDeg^¥circ$ to machine precision.

Under $¥varepsilon = ¥CondBPertEps$ perturbations to $C$
(¥CondBPertDraws{} random draws), the maximum principal angle across all
draws and all pairs is $¥CondBPertMaxRad$‾rad---well below the fragility
threshold of $¥CondBPertThreshold$‾rad, set at $1.5¥varepsilon$.
The threshold $1.5¥varepsilon$ is a heuristic calibration; the direct
numerical check confirms that the shared dominant eigenspace is robust to
this level of perturbation in $C$.
Structural coherence is perturbation-robust, distinguishing it from
accidental near-coincidences of eigenspaces that break under small
perturbations.

The numerical results are consistent with the theorem in all cases.

% -------------------------------------------------------------------------
¥section{Scope}
% -------------------------------------------------------------------------

¥paragraph{What is earned.}
In stationary finite-dimensional LGDS, among rank-$r$ linear observers
$¥phi_U(x)=U¥Sigma^{-1/2}x$ and squared-error prediction targets
$y=Cx_{t+¥tau}$, no common optimal rank-$r$ observer exists unless the
target family is $r$-coherent.
$r$-coherence requires all prediction targets to be functions of a common
$r$-dimensional sufficient statistic across all horizons---a strong alignment
between the system dynamics and the task family.
Outside $r$-coherence, no common optimal observer exists, and the cost of
applying any fixed observer to a non-native target is given exactly by the
Bayes risk formula.
In observer design workflows where a single rank-$r$ map must serve multiple
prediction channels, $¥Gamma$ is therefore a computable lower bound on
worst-case suboptimality under any shared observer: if $¥Gamma>0$, dedicated
per-target observers are necessary to achieve optimality.
The Grassmannian $¥Gr(r,n)$ is the literal domain of this optimisation:
a compact Riemannian manifold on which observer quality is a smooth function,
with $O(n)$ acting by transformation.
For fixed $C$, the computation reduces to a single $n ¥times n$
eigendecomposition of $M_{C,¥tau}$ per horizon once $A^¥tau$ is formed;
$A^¥tau$ is obtained in $O(n^3 ¥log ¥tau)$ operations by repeated squaring.

¥paragraph{What is not earned.}
This result concerns optimal reduced linear observation for prediction; it
does not address dynamic observer synthesis, asymptotic state reconstruction,
or feedback design.
It also does not apply to nonlinear systems, non-Gaussian noise,
nonlinear observers, or non-prediction tasks.
The breakdown is structural: for non-Gaussian noise or nonlinear dynamics,
the MMSE predictor is generally nonlinear and
Proposition‾¥ref{prop:opt}'s trace maximisation has no closed-form counterpart;
for nonlinear observers, the Grassmannian parametrisation and the Ky Fan
principle both cease to apply.
The causal axis---whether the observer optimal for intervention differs from
the observer optimal for prediction---is not addressed here.

¥paragraph{Extensions.}
The linear Gaussian setting admits closed-form analysis because the MMSE
predictor is linear, the stationary distribution is Gaussian, and the
optimal observer reduces to a single eigendecomposition.
Whether analogous shared-eigenspace conditions characterise common optimal
observers in non-Gaussian or nonlinear systems, and whether intervention
objectives impose additional constraints on observer selection, are open
questions beyond the scope of the present result.

% -------------------------------------------------------------------------
¥section*{Acknowledgements}
% -------------------------------------------------------------------------

The author received no external funding for this work.
Reproducible code is available from a publicly accessible repository
(details provided in the non-blind version of the manuscript).

¥bibliographystyle{elsarticle-num}
¥bibliography{refs}

¥end{document}
